---
title: 'Mapping Green Space and Traffic in Glasgow'
author: ~
date: '2017-11-10'
draft: no
slug: greentraffic
tags: ["Green space","Traffic", "Glasgow", "Neighbourhood","Rshiny"]
categories: ["Data Visualisation"]
---

## Background

The health benefit of greater availability of greenspaces is mediated by the quality of spaces [(Zhang et al., 2017)](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5451986/). Recently, quality was associated with psychological distress, whereas availability had no effect [(Feng and Astell-Burt, 2018)](https://bmcpsychiatry.biomedcentral.com/articles/10.1186/s12888-018-1926-1). Both of these measures of quality were self-reported. Another method to determine quality is from GIS-derived measures of land use. Perhaps the optimal method is using questionnaire self-report and GIS-derived measures as has been developed by [Stessens et al., 2017](https://www.sciencedirect.com/science/article/pii/S2212041616303898). Their measure of quality had the following subdomains and weightings: Naturalness and biodiversity (0.10), Spaciousness (0.16), Quietness (0.14) which were GIS-derived; cleanliness and maintenence (0.31), facilities (0.20) and feelings of safety (0.20) which were questionnaire-derived. In this work, I hypothesised that heavy traffic and accidents would influence quietness, and feelings of safety, thus reducing the quality of the greenspace. The first step then was to understand how traffic events and greenspaces covary, in Glasgow as a case study. 

## Research Questions
* Do traffic events occur close to greenspaces? 
* How does this vary temporally? 

## Method
I will describe the process of building a data visualisation to answer these questions. 
First load the packages:

```{r packages, eval=F, message=FALSE, warning=FALSE}
# allows r to work with spatial data
library(rgdal)
# allows r to connect with google apis
library(googleway)
# allows r to perform GIS operations (on vectors especially?)
library(rgeos)
# general purpose plotting package for r
library(ggplot2)
# download the greenspace map for scotland from:
# https://www.ordnancesurvey.co.uk/opendatadownload/products.html
```

Next, read in the green space data:

```{r greeendata, eval=F, message=FALSE, warning=FALSE}
# read the data
green<-readOGR("boundaries/green", "GB_GreenspaceSite")
# Download the local authority boundary for Glasgow City from here:
# https://saspac.org/data/2011-census/scotland-2011/
glasgow<-readOGR("boundaries/Scottish-Census-boundaries(shp)/2011_Council_Area/CA_2011_EoR_Glasgow_City", "CA_2011_EoR_Glasgow_City")
# clip to glasgow
green_subset <- green[glasgow, ]

# Project Glasgow greenspace (BNG) to lat/long (WGS84)
green_subset <- spTransform(green_subset, CRS("+init=epsg:4326"))

# Make it a dataframe
green_polygon<-fortify(green_subset)

# Tidy the coloumn names so that we can use it with the polyline encoding function
colnames(green_polygon)[1]<-"lon"
green_polygon<-subset(green_polygon, select=c("lon", "lat", "id"))

# Now for each group of points (vertices of polygon) we want to encode the polygon using the "encode_pl" function from the package googleway (comment block it do it doens't run everytime)

if(FALSE) {
polyline_df<-NULL
for (i in unique(green_polygon$id)){
dataG<-subset(green_polygon, id==i)
polyline<-encode_pl(dataG$lat, dataG$lon)
polyline_df<-rbind(df, polyline)
}
polyline_df<-as.data.frame(polyline_df)
colnames(polyline_df)<-"polyline"
}
}
# Let's write this out so that the end viz/application doesn't need to do this bit again
# write.csv(polyline_df, "data/polyline_df.csv")
```

Now, let's get historic traffic data:
```{r historictraffic, eval=F, message=FALSE, warning=FALSE}
# Data on historical traffic from here:
# https://www.dft.gov.uk/traffic-counts/area.php?region=Scotland&la=Glasgow+City
# "Briefly the file contains the annual traffic (otherwise known as volume of traffic) on each link of the major road network. This is calculated by multiplying the Annual average daily flow (AADF) by the corresponding length of road and by the number of days in the year""
data<-read.csv("data/Glasgow+City_TRAF.csv")
# make the data frame spatial
coords = cbind(data$Easting, data$Northing)
sp = SpatialPoints(coords)
# add the data
spdf = SpatialPointsDataFrame(coords, data)
# Set projection to bng
proj4string(spdf) <- CRS("+init=epsg:27700")
# project to bng
spdf <- spTransform(spdf, CRS("+init=epsg:4326"))
# Let's just look at all the motovehicles combineds
traffichistoric<-as.data.frame(spdf)
traffichistoric<-subset(traffichistoric, select=c("coords.x1", "coords.x2", "AllMotorVehicles"))
```

To get contemporary traffic information we can use the googleway package, i.e. current traffic volume on roads. All the data is then brought together in the Rshiny app:

```{r shinyapp, eval=F, message=FALSE, warning=FALSE}
# load packages
library(shiny)
library(googleway)
library(maptools)

# Read the processed data
traffichistoric<-read.csv("data/traffichistoric.csv")
polyline_df<-read.csv("data/polyline_df.csv")

# Create the shiny user input specs
ui <- fluidPage(google_mapOutput("map"))

# Create the shiny server specs
server <- function(input, output, session){

# Have to have an API key to use Google data  
map_key <- "Your-key-here"
  
# Style option not used but we can change the style of the map with the following code:
style <- '[{"featureType":"all","elementType":"all","stylers":[{"invert_lightness":true}, {"saturation":10},{"lightness":30},{"gamma":0.5},{"hue":"#435158"}]}, {"featureType":"road.arterial","elementType":"all","stylers":[{"visibility":"simplified"}]}, {"featureType":"transit.station","elementType":"labels.text","stylers":[{"visibility":"off"}]}]'
  
# output the map, use glasgow lat long to centre map, play with the zoom to make it look good
output$map <-renderGoogle_map({google_map(key = map_key, location=c(55.8642, -4.2518), zoom=13) %>%
# Add a heatmap of the historical traffic counts
add_heatmap(data = traffichistoric, lat = 'coords.x2', lon = 'coords.x1', option_opacity = 0.3) %>%
# This is the key bit to get contemporary traffic
add_traffic() %>% 
# And to the get the polygons of greenspace
add_polygons(data = polyline_df, polyline ="polyline", stroke_weight = 3, fill_colour = "green", stroke_colour = "black") 
})}
# run the app!
shinyApp(ui, server)
```

## Visualisation
Here is the final result:
```{r app, echo=F, message=FALSE, warning=FALSE}
library(knitr)
knitr::include_app("https://thebest.shinyapps.io/test_MC/")
```


  