---
title: 'Mapping Green Space and Traffic in Glasgow'
author: "Mark Cherrie"
date: '2017-11-10'
draft: no
slug: greentraffic
tags: ["Data visualisation", "Green space","Traffic", "Glasgow", "Neighbourhood", "Rshiny"]
categories: ["Data visualisations"]
---

The question, in relation to theories of how the quality of greenspace can affect relationships with health, might be: Do traffic events occur close to greenspaces? How does this vary spatially and temporally? How can this be visualised?

First load the packages:

```{r ui, eval=F, message=FALSE, warning=FALSE}
Load packages
# allows r to work with spatial data
library(rgdal)
# allows r to connect with google apis
library(googleway)
# allows r to perform GIS operations (on vectors especially?)
library(rgeos)
# general purpose plotting package for r
library(ggplot2)
# download the greenspace map for scotland from:
# https://www.ordnancesurvey.co.uk/opendatadownload/products.html
```

Next, read in the green space data:

```{r greeendata, eval=F, message=FALSE, warning=FALSE}
# read the data
green<-readOGR("boundaries/green", "GB_GreenspaceSite")
# Download the local authority boundary for Glasgow City from here:
# https://saspac.org/data/2011-census/scotland-2011/
glasgow<-readOGR("boundaries/Scottish-Census-boundaries(shp)/2011_Council_Area/CA_2011_EoR_Glasgow_City", "CA_2011_EoR_Glasgow_City")
## OGR data source with driver: ESRI Shapefile 
green_subset <- green[glasgow, ]

# Project Glasgow greenspace (BNG) to lat/long (WGS84)
green_subset <- spTransform(green_subset, CRS("+init=epsg:4326"))

# Make it a dataframe
green_polygon<-fortify(green_subset)

# Tidy the coloumn names so that we can use it with the polyline encoding function
colnames(green_polygon)[1]<-"lon"
green_polygon<-subset(green_polygon, select=c("lon", "lat", "id"))

# Now for each group of points (vertices of polygon) we want to encode the polygon using the "encode_pl" function from the package googleway (comment block it do it doens't run everytime)

if(FALSE) {
polyline_df<-NULL
for (i in unique(green_polygon$id)){
dataG<-subset(green_polygon, id==i)
polyline<-encode_pl(dataG$lat, dataG$lon)
polyline_df<-rbind(df, polyline)
}
polyline_df<-as.data.frame(polyline_df)
colnames(polyline_df)<-"polyline"
}
}
# Let's write this out so that the end viz/application doesn't need to do this bit again
# write.csv(polyline_df, "data/polyline_df.csv")
```

Now, let's get historic traffic data:
```{r historictraffic, eval=F, message=FALSE, warning=FALSE}
# Data on historical traffic from here:
# https://www.dft.gov.uk/traffic-counts/area.php?region=Scotland&la=Glasgow+City
# "Briefly the file contains the annual traffic (otherwise known as volume of traffic) on each link of the major road network. This is calculated by multiplying the Annual average daily flow (AADF) by the corresponding length of road and by the number of days in the year""
data<-read.csv("data/Glasgow+City_TRAF.csv")
# make the data frame spatial
coords = cbind(data$Easting, data$Northing)
sp = SpatialPoints(coords)
# add the data
spdf = SpatialPointsDataFrame(coords, data)
# Set projection to bng
proj4string(spdf) <- CRS("+init=epsg:27700")
# project to bng
spdf <- spTransform(spdf, CRS("+init=epsg:4326"))
# Let's just look at all the motovehicles combineds
traffichistoric<-as.data.frame(spdf)
traffichistoric<-subset(traffichistoric, select=c("coords.x1", "coords.x2", "AllMotorVehicles"))
```

We can get the relevant information from the googleway package, i.e. current traffic volume. So all that's needed is to put it in a shiny app for the final data visualisation:

```{r shinyapp, eval=F, message=FALSE, warning=FALSE}
# load packages
library(shiny)
library(googleway)
library(maptools)

# Read the processed data
traffichistoric<-read.csv("data/traffichistoric.csv")
polyline_df<-read.csv("data/polyline_df.csv")

# Create the shiny user input specs
ui <- fluidPage(google_mapOutput("map"))

# Create the shiny server specs
server <- function(input, output, session){

# Have to have an API key to use Google data  
map_key <- "Your-key-here"
  
# Style option not used but we can change the style of the map with the following code:
style <- '[{"featureType":"all","elementType":"all","stylers":[{"invert_lightness":true}, {"saturation":10},{"lightness":30},{"gamma":0.5},{"hue":"#435158"}]}, {"featureType":"road.arterial","elementType":"all","stylers":[{"visibility":"simplified"}]}, {"featureType":"transit.station","elementType":"labels.text","stylers":[{"visibility":"off"}]}]'
  
# output the map, use glasgow lat long to centre map, play with the zoom to make it look good
output$map <-renderGoogle_map({google_map(key = map_key, location=c(55.8642, -4.2518), zoom=13) %>%
# Add a heatmap of the historical traffic counts
add_heatmap(data = traffichistoric, lat = 'coords.x2', lon = 'coords.x1', option_opacity = 0.3) %>%
# This is the key bit to get contemporary traffic
add_traffic() %>% 
# And to the get the polygons of greenspace
add_polygons(data = polyline_df, polyline ="polyline", stroke_weight = 3, fill_colour = "green", stroke_colour = "black") 
})}
# run the app!
shinyApp(ui, server)
```

```{r app, echo=F, message=FALSE, warning=FALSE}
library(knitr)
knitr::include_app("https://thebest.shinyapps.io/test_MC/",height ="600px")
```
  