---
title: What are the optional courses like for UoE MSc students?
author: ~
date: '2018-05-22'
slug: MscCourses
categories: ["Data Processing"]
tags: ["R", "Web Scraping", "Education", "Admin"]
draft: no
---
The aim was to understand more about the optional courses available for MSc students in the school of Geosciences. This meant building a dataframe that contained the interesting bits of information available on the University of Edinburgh website so that I could query it to find out answers to questions such as: How many 10 credit courses are available? Which degree cluster do the courses belong to and is there any overlap? What kind of topics/skills are going to be taught? and so on. 

The first thing to do was to download the data from the University of Edinburgh website. Here is the code, which first builds the URL's that we need and then scrapes the relevant info one-by-one into a dataframe:


```{r collection, echo=TRUE, message=FALSE}
############### DATA COLLECTION #########################

## scaping the edinburgh university website
library(rvest)
# Gather the optional courses for Earth Observation and Geoinformation Management
MScOptionsURLS<-function(MSc){
if(MSc=="EOGM"){
  MSc<-"https://www.ed.ac.uk/geosciences/postgraduate/taught-masters/msc-earth-observation/structure-teaching-assessment/optional-courses"
  url<-MSc %>%
  read_html()%>%
  html_nodes('#block-system-main p:nth-child(1) a , p+ p a')%>%
  html_attr('href')
  } else if(MSc=="GIS"){
  MSc<-"https://www.ed.ac.uk/geosciences/postgraduate/taught-masters/geographical-information-science/structure-teaching-assessment/optional-courses"
  url<-MSc %>%
    read_html()%>%
    html_nodes('#block-system-main p:nth-child(1) a , p+ p a')%>%
    html_attr('href')
  }else if(MSc=="ARCH"){
  MSc<-"https://www.ed.ac.uk/geosciences/postgraduate/taught-masters/msc-gis-and-archaeology/structure-teaching-and-assessment/optional-courses"
  url<-MSc %>%
    read_html()%>%
    html_nodes('.col-sm-9 , p+ p a')%>%
    html_attr('href')
  # not perfect, but get the dpt links
  url<-url[3:10]
  # weird, not picking up the final two, must be formatted differently, quick fix
  url[9]<-"http://www.drps.ed.ac.uk/17-18/dpt/cxpghc11412.htm"
  url[10]<-"http://www.drps.ed.ac.uk/17-18/dpt/cxpghc11075.htm"
  }else {
    "Pick a MSc course that's in the Spatial Informatics and Earth Observation cluster"
  }
url<-as.data.frame(url)
colnames(url)<-"url"
url$url<-as.character(url$url)
return(url)
}

# Let's do it for the 3 MSc courses
EOGM_options<-MScOptionsURLS(MSc = "EOGM")
GIS_options<-MScOptionsURLS(MSc = "GIS")
ARCH_options<-MScOptionsURLS(MSc = "ARCH")

# Now let's scrape the optional courses pages
eduniscraper<-function(url){
  Urlcatch<-tryCatch({
    # Name of course
    Name<-tryCatch({
      Name<-url %>%
        read_html() %>%
        html_node("#sitspagetitle") %>%
        html_text()
     #Name<-gsub("Undergraduate Course:|Postgraduate Course: ", "", Name)
    }, error=function(e){
      Name<-NA}
    )
    
    # Get the whole table
    tableinfo<-url %>%
      read_html() %>%
      html_nodes('.sitstablegrid') %>%
      html_text()
    
    ## Credits
    SCQFcredits<-tryCatch({
      subtable<-tableinfo[gregexpr(pattern ='SCQF Credits',tableinfo)>1]
      start<-gregexpr(pattern ='SCQF Credits',subtable)[[1]][1]+12
      end<-gregexpr(pattern ='SCQF Credits',subtable)[[1]][1]+13
      SCQFcredits<-as.numeric(substring(subtable, start,end))
    }, error=function(e){
      SCQFcredits<-NA}
    )
    
    ## Time
    Time<-tryCatch({
      subtable<-tableinfo[gregexpr(pattern ='Total Hours',tableinfo)>1]
      start<-gregexpr(pattern ='Total Hours',subtable)[[1]][1]
      end<-gregexpr(pattern ='Assessment \\(Further Info\\)',subtable)[[1]][1]-1
      Time<-substring(subtable, start,end)
      Time<-gsub("[\n]|:|\\)|\\(\\)|W|Assessment|(Further Info)|", "", Time)
      Time<-gsub("\\(", ",", Time)
    }, error=function(e){
      Time<-NA}
    )
    
    ## Assessment
    assessment<-tryCatch({
      subtable<-tableinfo[gregexpr(pattern ='Written Exam',tableinfo)>1]
      start<-gregexpr(pattern ='Written Exam',subtable)[[1]][1]
      end<-start+55
      assessment<-substring(subtable, start,end)
      assessment<-gsub(" |[\t]|-|[\n]", "", assessment)
      assessment<-gsub("WrittenExam|Coursework|PracticalExam", "", assessment)
      assessment<-gsub("%", "", assessment)
      WrittenExam=as.numeric(unlist(strsplit(assessment, ","))[1])
      Coursework=as.numeric(unlist(strsplit(assessment, ","))[2])
      PracticalExam=as.numeric(unlist(strsplit(assessment, ","))[3])
    }, error=function(e){
      WrittenExam<-NA
      Coursework<-NA
      PracticalExam<-NA}
    )
    
    # learning outcomes
    learningoutcomes<-tryCatch({
      learningoutcomes<-tableinfo[gregexpr(pattern ='By the end of this course students will be able to|On completion of this course|On completion of the course|Students will be able to:',tableinfo)>1][2]
      #if (is.na(learningoutcomes)){
      #learningoutcomes<-tableinfo[gregexpr(pattern ='On completion of this course|On completion of the course|Students will be able to:',tableinfo)>1][1]
      #}
      learningoutcomes<-gsub("Learning Outcomes|[\t]|-|[\n]", "", learningoutcomes)
    }, error=function(e){
      learningoutcomes<-NA}
    )
    
    ## Skills
    skills<-tryCatch({
      subtable<-tableinfo[gregexpr(pattern ='Graduate Attributes and Skills',tableinfo)>1][2]
      start<-gregexpr(pattern ='Graduate Attributes and Skills',subtable)[[1]][1]
      end<-gregexpr(pattern ='Special Arrangements|Keywords|Additional Class Delivery Information',subtable)[[1]][1]-1
      skills<-substring(subtable, start,end)
      skills<-gsub("Graduate Attributes and Skills|[\t]|-|[\n]", "", skills)
    }, error=function(e){
      skills<-NA}
    )
    
    df<-data.frame(Name, url, SCQFcredits, Time, learningoutcomes, skills, WrittenExam, Coursework, PracticalExam)
    #write.csv(df, paste0("data/", url, ".csv"), row.names=F)
    return(df)
  },error=function(e){
    df<-data.frame(Name=NA, url=url, SCQFcredits=NA, Time=NA, learningoutcomes=NA, skills=NA, WrittenExam=NA, Coursework=NA, PracticalExam=NA)}
  )
}

# test
url="http://www.drps.ed.ac.uk/17-18/dpt/cxpghc11412.htm"
test<-eduniscraper(url)

# loop
library(plyr)
# Technologies for Sustainable Energy, i.e. number 6, doesn't work
#EOGM_options <- subset(EOGM_options, url!= "http://www.drps.ed.ac.uk/16-17/dpt/cxpgee10001.htm")
#Put in error handling instead
EOGM_df<-mdply(EOGM_options, eduniscraper)
GIS_df<-mdply(GIS_options, eduniscraper)
ARCH_df<-mdply(ARCH_options, eduniscraper)
```

Next step was to apply some cleaning, this was a iterative process of identifying issues with the output above and applying correction. Have a look at the code comments for more details:

```{r cleaning, echo=TRUE, message=FALSE, warning=FALSE}

############### DATA CLEANING #########################

# master data
EOGM_df$cluster<-"EOGM"
GIS_df$cluster<-"GIS"
ARCH_df$cluster<-"ARCH"
master_df<-rbind(EOGM_df, GIS_df, ARCH_df)

# 19 courses are duplicates
library(dplyr)
master_df<-master_df %>% 
  group_by(url) %>% 
  mutate(cluster_group = paste0(cluster, collapse = ",")) 
master_df<-master_df[!duplicated(master_df$Name),]
master_df$cluster<-NULL

# 3 course links don't work, but only one remains,
# as it is considered 'unique in above
master_df<-master_df[!is.na(master_df$Name),]


# Code NA to not entered
master_df[is.na(master_df)]<-"Not Entered"

# postgrad or undergrad
master_df$coursetype<-ifelse(grepl("Undergraduate", master_df$Name), "U", "P")
master_df$Name<-gsub("Undergraduate Course:|Postgraduate Course:", "", master_df$Name)

# get time vars
result<-list()

for(i in 1:33){
Time=unlist(strsplit(master_df$Time[i], ","))
Time<-gsub(" ", "", Time)
Time<-as.data.frame(Time)
colnames(Time)<-"TimeVars"
library(tidyr)
Time<-Time %>%
separate(TimeVars, 
         into = c("text", "num"), 
         sep = "(?<=[A-Za-z])(?=[0-9])") 
Time$Name<-master_df$Name[i]
result[[i]] <- Time
}
timevars = do.call(rbind.fill, result)
timevars = timevars[!is.na(timevars$text),]
timevars<-timevars %>%
  spread(text, num)
timevars[is.na(timevars)]<-0
timevars<-subset(timevars, Name!=' Water Resource Management (PGGE11018)')
timevars<-subset(timevars, Name!=' Epidemiology for Public Health (PUHR11016)')

# merge with master df
master_df<-merge(master_df, timevars, by="Name", all.x=T)
master_df$Time<-NULL
master_df$NotEntered<-NULL
```

Finally, onto the fun part, querying the data and producing graphs to present answers to the questions I was interested in. First let's see the proportion of courses that are 10 and 20 credits:

```{r visualisation1, echo=F, message=FALSE, warning=FALSE}
# credits
library(ggplot2)
ggplot(master_df) + stat_count(mapping = aes(x=as.factor(SCQFcredits), y=..prop.., group=1))
summary(master_df$SCQFcredits)
```

* So we can see that 1/3 of the optional courses are currently 10 credits

Next, I want to see how they relate to the degree clusters:

```{r visualisation2, echo=F, message=FALSE, warning=FALSE}
# cluster group
library(ggplot2)
g<- ggplot(master_df) + aes(cluster_group)
g + geom_bar()
```

* EOGM has the most unique master courses (10)
* EOGM and GIS share 11 courses
* There is only one course listed in all of the masters (Principles and Practice of Remote Sensing (PGGE11233))

Next, I'm interested in the assessment breakdown (coursework, written exam and practical exam):

```{r visualisation3, echo=F, message=FALSE, warning=FALSE}
# Coursework
master_df$assessmentbreakdown<-paste0("C", master_df$Coursework, 
                                      "W", master_df$WrittenExam, 
                                      "P", master_df$PracticalExam)
ggplot(master_df) + stat_count(mapping = aes(x=assessmentbreakdown, y=..prop.., group=1))
table(master_df$assessmentbreakdown)
```

* Two thirds of the courses are assessed via coursework only
* The second most common option is a 50/50 split between coursework and a Written exam


Finally, let's have a look at the learning outcomes, which will give me an understanding of the topics and skills that the student will have experience of by the end: 

```{r visualisation4, echo=F, message=FALSE, warning=FALSE}
### topic modelling for the skills and learning outcomes
library(topicmodels)
library(tm)
master_df$learningoutcomes<-gsub("On completion of this course, the student will be able to:|By the end of this course students will be able to:|Students will be able to:|On completion of the course, a student should", "", master_df$learningoutcomes)
text_corpus <- VCorpus(VectorSource(master_df$learningoutcomes))

# clean up
text_corpus_clean <- tm_map(text_corpus,
                            content_transformer(tolower))
#text_corpus_clean <- tm_map(text_corpus_clean, stemDocument)
text_corpus_clean <- tm_map(text_corpus_clean, removeNumbers)
text_corpus_clean <- tm_map(text_corpus, removeWords, c(stopwords("english"),"understand","understanding",
                                                        "demonstrate","including",
                                                        "different", "able", 
                                                        "students", "ability", "coursework",
                                                        "examination", "have", "use"
                                                        ,"key", "group")) 
text_corpus_clean <- tm_map(text_corpus_clean, removePunctuation)
text_corpus_clean <- tm_map(text_corpus_clean, stripWhitespace)

# word cloud
library(wordcloud)
wordcloud(text_corpus_clean, min.freq = 2, random.order = FALSE,
          colors=brewer.pal(8, "Dark2"))
```

There is probably a lot more that can be done with the data but this is a just a taster. In addition, this method could be applied to other schools (not just geosciences).



